p-adic stability in linear algebra

### 2015-01-18 examples/stability.tex
Update charpoly section

Such polygons arise naturally in tracking the precision of polynomials 
\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose 
Newton polygons lie above $P$.  This lattice is generated by monomials 
$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
One uses these polygons to do a certain kind of coordinate-wise 
precision tracking for polynomials, bounding the actual precision 
lattice. We now introduce another polygon that will be bounded by 
between these two, and will provide an estimate for $\mathcal{E}$.

@@

It is clear from the definition $\mathcal L_{\PP(M)} \subset 
\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
$M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
Proposition \ref{prop:precision}, the precision polygon determines the 
best precision loses we can hope if we are encoding precision on 
polynomials by polygons.

It turns out that the precision polygon is related to Hodge and Newton 
polygon. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
the translated polygon with vertices $(x_i - n, y_i)$.

@@

It remains to prove the comparison with Newton polygon.
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$, 
$f_{i,j}$ as the $(i,j)$th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
\val(m_{i,j})$. We write $f[k]$ 
for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
for all $i$ and $k$,
\begin{align*}
f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
\mu_{i,n} + f_{n,i}[k]) \\
&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
\end{align*}
with the infimum is taken over $j$. 
Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing
(it follows from the comparison with the Hodge polygon), we get the 
result.
\end{proof}

\begin{rem}
Experiments actually support the following stronger result: $\PP(M)$ is 
bounded above by $T_1(\HP(M))$.
\end{rem}

@@

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
using the Hodge polygon loses precision but not so much. Nonetheless, as 
already illustrated for matrix multiplication, this loses of precision 
may accumulate and become significant if we need to compute many 
characteristic polynomials. In the case of a single characteristic 
polynomial, we note moreover that if the $\sigma_i(M)$'s are known to be 
large or if $M$ is a translate of such a matrix, using lattice precision 
can be very useful.

\begin{comment}
\subsection{Characteristic polynomials} \label{ssec:charpoly}

Write $\charpoly : M_{n,n}(K) \to K[X]$ for the characteristic polynomial, and $K_{<n}[X] \subset K[X]$ for the subspace consisting of polynomials of degree less than $n$.  Then the determinant of $\charpoly$ at a point $M$ is
\[
\charp(M) : dM \mapsto \tr(\com(X - M) \cdot dM).
\]
The image is the $K$-span of the entries of $\com(X - M)$, which is clearly contained within $K_{<n}[X]$.  In fact, the image will equal $K_{<n}[X]$ as long as $M$ does not have two Jordan blocks with the same generalized eigenvalue.

Write $\mathcal{E}$ for the precision lattice in $K_{<n}[X]$ of the characteristic polynomial of $M$, where $M$ is known with precision lattice $\mathcal{L}$.  As long as $\mathcal{L}$ is sufficiently small, Proposition \ref{prop:precision} implies that $\mathcal{E}$ will be generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $\mathcal{L}$.  When $\mathcal{L} = \mathcal{L}_N := \pi^N M_{n,n}(\O_K)$, we can further analyze $\mathcal{E}$.

Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = \sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is $\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$.  Such polygons arise naturally in tracking the precision of polynomials \cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose Newton polygons lie above $P$.  This lattice is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  One uses these polygons to do a certain kind of coordinate-wise precision tracking for polynomials, bounding the actual precision lattice.

For any matrix $M$, the Newton polygon lies above the Hodge polygon \cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.  We now introduce another polygon that will be bounded by between these two, and will provide an estimate for $\mathcal{E}$.

\begin{deftn}
The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
\end{deftn}

If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be the translated polygon with vertices $(x_i - n, y_i)$ and $S_n(P)$ be the translated polygon with vertices $(x_i, y_i + n)$.

\begin{prop}
Suppose a matrix $M$ is known with precision $\mathcal{L}_N$ with $N$ sufficiently large.  Then $\mathcal{E} \subseteq \mathcal{L}_{S_N(\PP(M))}$.
\end{prop}
\begin{proof}
The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(\mathcal{L}_0)$, the result follows from Proposition \ref{prop:precision}.
\end{proof}

\begin{prop} \label{prop:prec_hodge}
The precision polygon $\PP(M)$ is bounded below by the translated Hodge polygon $T_1(\HP(M))$.
\end{prop}

\begin{proof}
The coefficients of $\charpoly(M)$ can be expressed as traces of exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M)) = \tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.  The entries of $\Lambda^i(\Delta_M)$ can be computed easily, and now the result follows from Proposition \ref{prop:precision2}. % Clarify and expand.
\end{proof}

For many matrices, $\PP(M) = T_1(\HP(M))$.  For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  For $p=3$, the fraction rises to $99.98\%$, with no clear dependence on dimension.

We may also bound $\PP(M)$ above:

\begin{prop} \label{prop:newton_prec}
If $M \in M_{n,n}(\O_K)$, the precision polygon $\PP(M)$ is bounded above by the Newton polygon $\NP(M)$.
\end{prop}
\begin{proof}
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$ and $f_{i,j}$ as the $(i,j)$th entry of $\com(X - M)$. We write $f[k]$ for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = +\infty$. The equation $(X - M) \cdot \com(X - M) = f \cdot I$  yields, for all $i$ and $k$,
\begin{align*}
f[k] &= \inf(f_{i,i}[k-1], \val(m_{i,0}) + f_{0,i}[k], \ldots, \val(m_{i,n}) + f_{n,i}[k]) \\
&\ge \inf(f_{j,i}[k]),
\end{align*}
with the infimum taken over $j$.  Taking lower convex hulls yields the result.
\end{proof}

Moreover, experiments support the following stronger results.

\begin{conj}
\begin{enumerate}
\item $\PP(M) = T_1(\HP(M))$ if $\val(\det(M))$ is $0$ or $1$, or if $M \in M_{3,3}(\O_K)$.
\item The precision polygon $\PP(M)$ is bounded above by $T_1(\NP(M))$.
\end{enumerate}
\end{conj}
Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at $1$, corresponding to the precision of the linear term of the characteristic polynomial.

Of course, $\PP(M)$ does necessarily agree with $\mathcal{E}$.  In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the index of $\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.  This percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.

Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : \mathcal{E}]$ arbitrarily large.  Suppose the $\sigma_i(M)$ are large.  Proposition \ref{prop:prec_hodge} implies that $\mathcal{E}$ is contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} \sigma_i(M)$.  The precision lattice of $1 + M$ is obtained from $\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ is now flat with height $0$.

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ using the Hodge polygon rarely loses much precision.  However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.
\end{comment}

>>>>

Such polygons arise naturally in tracking the precision of polynomials 
\cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a 
lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose 
Newton polygons lie above $P$.  This lattice is generated by monomials 
$a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  
One uses these polygons to do a certain kind of coordinate-wise 
precision tracking for polynomials, bounding the actual precision 
lattice. We now introduce another polygon, bounded 
between these two, that will provide an estimate on the precision of $\charpoly(M)$.

@@

It is clear from the definition $\mathcal L_{\PP(M)} \subset 
\charp(M)(\mathcal{L}_0)$ where $\mathcal{L}_0$ is the standard lattice 
$M_{n,n}{\OK}$. More precisely $\PP(M)$ is the smallest polygon $P$ for 
which the inclusion $\mathcal L_P \subset \charp(M)(\mathcal{L}_0)$. By 
Proposition \ref{prop:precision}, the precision polygon determines the minimal 
precision losses possible when encoding polynomial precision using polygons.

It turns out that the precision polygon is related to the Hodge and Newton 
polygons. If a polygon $P$ has vertices $(x_i, y_i)$, we let $T_n(P)$ be 
the translated polygon with vertices $(x_i - n, y_i)$.

@@

It remains to prove the comparison with Newton polygon.
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$, 
$f_{i,j}$ as the $(i,j)$th entry of $\com(X{-}M)$ and $\mu_{i,j} = 
\val(m_{i,j})$. We write $f[k]$ 
for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = 
+\infty$. The equation $(X{-}M) \cdot \com(X{-}M) = f \cdot I$ yields, 
for all $i$ and $k$,
\begin{align*}
f[k] &\ge \inf(f_{i,i}[k{-}1], \mu_{i,0} + f_{0,i}[k], \ldots, 
\mu_{i,n} + f_{n,i}[k]) \\
&\ge \inf(f_{i,i}[k{-}1], f_{j,i}[k]),
\end{align*}
with the infimum taken over $j$. 
Taking lower convex hulls and noting that $\PP(M)$ is nonincreasing
(which follows from the comparison with the Hodge polygon), we get the 
result.
\end{proof}

\begin{rem}
Experiments actually support the following stronger result: $\PP(M)$ is 
bounded above by $T_1(\NP(M))$.
\end{rem}

@@

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ 
using the Hodge polygon loses only a small amount of precision. However,
if the $\sigma_i(M)$'s are known to be 
large or if $M$ is a translate of such a matrix, using lattice precision 
can be very useful.
